Linear stability of magnetized massive protoplanetary disks

### 2014-04-02 paper/results.tex
tex updates

Fig. \ref{mri_massive_resis} plots the energies assocaited with the 
MRI mode discussed above. 
%with $k_xH\sim1.3$ in the $\Lambda_0=0.1$  
%disk. 
The gravitational energy exceeds the 
magnetic energy near the midplane ($z\lesssim0.2H$). The growth rate $\gamma=0.25\Omega$
is not much smaller than that of the most unstable mode
($\gamma=0.36\Omega$ for $k_xH=0.1$), so significant density
perturbations are expected to grow on dynamical timescales for this
system, even though GI should not develop.

@@

The above assumptions imply
\begin{align}
  W^{(1)} \sim \frac{\Omega^2}{c_s^2 Q k^2} W^{(0)},\label{feedback}
\end{align} 
which indicates a non-zero density perturbation due to the
MRI can be amplified by self-gravity. Now, for $k_xH\sim 1$ we have
$|k_z^2/k_x^2|\sim \beta/f\gg1$ because $f=O(1)$ and $\beta\gtrsim10$
for the cases considered above. Then
\begin{align}
  \left|\frac{W^{(1)}}{W^{(0)}}\right| \sim \frac{1}{Q\beta}, 
\end{align}
 suggesting stronger amplification of the density field by 
 self-gravity with increasing field strength (decreasing
 $\beta$).

@@

 The above arguments can be adapted to the resistive
 limit. Eq. \ref{simp3}---\ref{simp4} are unchanged, while resistive
 terms appearing in Eq. \ref{simp1}---\ref{simp2} only involve the
 potential perturbation through $\w$. For the resistive MRI we assume
 $k_z\sim v_A/\eta$ and $|\sigma|\sim v_A^2/\eta = \Lambda\Omega$
 \citep{sano99}. Then $|\sigma^2/c_s^2k_z^2|\sim 1/\beta \ll 1$ so
 Eq. \ref{cond} is satisfied. Noting that $k_z^2\sim
 \Lambda^2\Omega^2\beta/c_s^2$, the feedback equation becomes
 \begin{align}
   \left|\frac{W^{(1)}}{W^{(0)}}\right| \sim
   \frac{1}{Q\left(f^2\hat{k}_x^2 + \beta\Lambda^2\right)},
 \end{align}
 so increasing the resistivity (decreasing $\Lambda$) should enhance
 density perturbations.

 For weak fields, the MRI has small vertical wavelength,
 implying rapid density variations with height. The magnitude of the
 corresponding gravitational potential is then expected to be small
 because of internal cancellations, since the potential is essentially
 the integral of the density field.

>>>>

Fig. \ref{mri_massive_resis} plots the energies associated with the 
MRI mode discussed above. 
%with $k_xH\sim1.3$ in the $\Lambda_0=0.1$  
%disk. 
The gravitational energy exceeds the 
magnetic energy near the midplane ($z\lesssim0.2H$). The growth rate $\gamma=0.25\Omega$
is not much smaller than that of the most unstable mode
($\gamma=0.36\Omega$ for $k_xH=0.1$), so significant density
perturbations are expected to grow on dynamical timescales for this
system, even though GI should not develop.

@@

The above assumptions imply
\begin{align}
  W^{(1)} \sim \frac{\Omega^2}{c_s^2 Q k^2} W^{(0)},\label{feedback}
\end{align} 
which indicates a non-zero density perturbation due to the
MRI can be amplified by self-gravity. Now, for $k_xH\sim 1$ we have
$|k_z^2/k_x^2|\sim \beta/f^2\gg1$ because $f=O(1)$ and $\beta\gtrsim10$
for the cases considered above. Then
\begin{align}
  \left|\frac{W^{(1)}}{W^{(0)}}\right| \sim \frac{1}{Q\beta}, 
\end{align}
 suggesting stronger amplification of the density field by 
 self-gravity with increasing field strength (decreasing
 $\beta$).

@@

 The above arguments can be adapted to the resistive
 disk. Eq. \ref{simp3}---\ref{simp4} are unchanged, while resistive
 terms appearing in Eq. \ref{simp1}---\ref{simp2} only involve the
 potential perturbation through $\w$. For the resistive MRI we assume
 $k_z\sim v_A/\eta$ and $|\sigma|\sim v_A^2/\eta = \Lambda\Omega$
 \citep{sano99}. Then $|\sigma^2/c_s^2k_z^2|\sim 1/\beta \ll 1$ so
 Eq. \ref{cond} is satisfied. Noting that $k_z^2\sim
 \Lambda^2\Omega^2\beta/c_s^2$, the feedback equation becomes
 \begin{align}
   \left|\frac{W^{(1)}}{W^{(0)}}\right| \sim
   \frac{1}{Q\left(f^2\hat{k}_x^2 + \beta\Lambda^2\right)},
 \end{align}
 so increasing the resistivity (decreasing $\Lambda$) should enhance
 density perturbations.

 %mass contained in small wavelength is small

 For weak fields in an ideal disk, the MRI has a small vertical
 wavelength, $\lambda \ll H$. The perturbed mass contained within
 $\sim \lambda$ is small and its potential
 perturbation is unimportant. Furthermore, considering the stratified
 disk, $\lambda\ll H$ imply rapid variations in the density
 perturbation across the disk height, averaging to zero, so the
 magnitude of the associated potential perturbation is small. 
% because the gravitational potential is the integral of the density
% field, i
 Self-gravity plays does not affect the MRI in this regime.

 %, i.e. internal cancellations would occur when such an integral is
 % performed.

 However, a strong field and/or large resistivity increases the MRI
 vertical wavelength. When the vertical scale of the MRI becomes
 comparable to the disk thickness, i.e. $\lambda\sim H$, the
 perturbed mass across the disk height can contribute to a net potential
 perturbation. We therefore expect self-gravity to be important for
 the MRI when it is weak.

%The perturbed mass
% within the unstable mode corresponds to that in the within 
%This can give rise to a net
% potential perturbation  
 %Self-gravity can then be effective, since it
 % operates globally in the vertical direction. 
%GI of 3D disks can be
 %analysed in vertically-integrated
 %systems 
%coupling to SG 
%That is when the MRI is weak

### 2014-04-02 paper/results_azi.tex
tex updates

>>>>

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_resis_ep0.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0.52cm]{figures/compare_growth3_tilted_resis_ep1.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_resis_ep2.ps}
  \caption{Growth rates in isothermal resistive disks with $Q=0.18$
    ($Q_\mathrm{2D}=0.67$) and $\Lambda_0=0.1$.  
    \label{compare_growth3_tilted_resis}}
\end{figure}

### 2014-04-02 paper/results_gi.tex
tex updates

\section{Gravitationally unstable disks} \label{result2}
Gravitational instability (GI) becomes possible in a sufficiently
massive and/or cold disk. We explore how the emergence of GI affects
the MRI, and vice versa. To do so, we compute unstbale modes for  
isothermal disks with $Q < 0.2 $ ($Q_\mathrm{2D}\lesssim 0.67$) which
permits gravitational instability as well as the MRI.

\subsection{Co-existence of MRI and GI}
We first consider ideal MHD by setting $\Lambda_0=100$ and $A=1$. 
Fig. \ref{compare_growth3} show growth rates for 
modes with $k_xH=1$ as a function of $\beta$ in disks with $Q=0.18$,
$Q=0.14$ and $Q=0.12$.

All three cases display distinct GI modes (red/brown branch). The GI
growth rates are $\gamma\simeq 0.25\Omega,\,0.6\Omega,\,0.8\Omega$ for
$Q=0.18,\,0.14,\,0.12$, respecitvely. GI is stabilized by magnetic
pressure for sufficiently small $\beta$, and the critical field
strength increases with the self-gravity \citep{nakamura83}. For
$Q=0.18$, GI is stabilized for $\beta \lesssim 15$. Nevertheless, 
the MRI branch for $\beta < 15$ becomes self-gravitating, so that
density perturbations are still expected even though  GI does not
formally operate.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0cm]{figures/compare_growth3_kx1_Q0d18.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d14.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d12.ps} 
  \caption{Growth rates for linear modes with $k_xH=1$ in an isothermal disk with
    $Q=0.18$ (top), $Q=0.14$ (middle) and $Q=0.14$ (bottom). The
    colorbar measures the importance of 
    self-gravity.  
    \label{compare_growth3}}
\end{figure}

The GI and MRI branches only interact when their growth rates
are similar. This is seen in Fig. \ref{compare_growth3} for $Q=0.18$
where the GI branch approaches an MRI branch at $\beta\simeq 25,\,
\gamma\simeq 0.2\Omega$. In fact following the red curve to smaller
$\beta$ indicates GI transitions to MRI. The
`gaps' in the GI and MRI branches for $Q=0.18$ and $Q=0.12$ may be due
to avoided crossing of modes, as seen in stars
\citep[e.g.][]{aizenman77} and accretion tori 
\citep[e.g.][]{christo93}. However, we cannot exclude the possibility 
that some modes may have been missed in the numerical search of
eigenfrequencies.

Thus, our results do not rigorously prove that the GI and MRI branches
do not intersect.  Nevertheless, the continous variation of $\tau$ with
respect to $\beta$ strongly suggest that unstable modes can transition
smoothly from magnetically-dominated to gravitationally-dominated at
low $\beta$.

\subsubsection{Case study} 
In reality, perturbations with a range of $k_x$ will be present for a
given set of disk parameters. Fig. \ref{compare_growth3_Q01d2} show
growth rates as a function of $k_x$ for an isothermal disk with
$Q=0.12,\,\beta=20$. MRI and GI are active with comparable growth
rates. All perturbations with $k_xH \lesssim 3.5$ grow dynamically
($\gamma\gtrsim 0.1\Omega$, or $\lesssim$ 1.6 orbits).

@@

The dispersion relation $\gamma(k_x)$ increasingly deviates from the
pure GI/MRI curves with $k_x$, implying increasing interaction between
magnetic and density perturbations. 
Comparing pure GI (dashed line)  and the gravitationally-dominated
portions of $\gamma(k_x)$ indicate inclusion of magnetic field
stabilizes high-$k_x$ pure GI. (Note also the slight decrease in the most unstable
$k_x$.) Since axisymmetric GI is unaffected by shear, this
stabilization is due to magnetic pressure and is consistent with
\citep{lizano10}.

@@

Comparing pure MRI (solid line) and the magnetically-dominated
portions of $\gamma(k_x)$ show that self-gravity increases MRI growth
rates at large $k_x$, but the effect is small. Note that this
destabilization is through the linear response, rather than through
the background stratification.

@@

>>>>

\section{Gravitationally unstable disks} \label{result2}
Gravitational instability becomes possible in a sufficiently
massive and/or cold disk. Here, we explore whether or not GI and MRI
can interact by computing unstable modes for isothermal disks with $Q < 0.2 $
($Q_\mathrm{2D}\lesssim ~0.67$) which permits GI, as shown below.
We consider ideal disks with $\Lambda_0=100$ and $A=1$, unless
otherwise stated.

\subsection{Co-existence of MRI and GI} 
Fig. \ref{compare_growth3} show growth rates for modes with $k_xH=1$
as a function of $\beta$ in disks with $Q=0.18$, $Q=0.14$ and
$Q=0.12$.   
All three cases display distinct GI modes (red/brown branch). The GI 
growth rates are $\gamma\simeq 0.25\Omega,\,0.6\Omega,\,0.8\Omega$ for
$Q=0.18,\,0.14,\,0.12$, respecitvely. GI is stabilized by magnetic
pressure for sufficiently small $\beta$, and the critical
field strength increases with increasing self-gravity \citep{nakamura83}. 
For $Q=0.18$, GI is stabilized for $\beta \lesssim 15$. Nevertheless, 
the MRI branch for $\beta < 15$ becomes self-gravitating, so that
density perturbations will still develop, even though GI does not
formally operate.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0cm]{figures/compare_growth3_kx1_Q0d18.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2.1cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d14.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.5cm]{figures/compare_growth3_kx1_Q0d12.ps} 
  \caption{Growth rates for modes with $k_xH=1$ in an isothermal disk with
    $Q=0.18$ (top), $Q=0.14$ (middle) and $Q=0.14$ (bottom). The
    colorbar measures the importance of  
    self-gravity with $\tau$.   
    \label{compare_growth3}}
\end{figure}

The GI and MRI branches only interact when their growth rates
are similar. This is seen in Fig. \ref{compare_growth3} for $Q=0.18$
where the GI branch approaches an MRI branch at $\beta\simeq 25,\,
\gamma\simeq 0.2\Omega$. In fact, following the red curve to smaller
$\beta$ indicates GI transitions to MRI. The
`gaps' in the GI and MRI branches for $Q=0.18$ and $Q=0.12$ may be due
to the phenomenon of avoided crossing, as seen in stars
\citep[e.g.][]{aizenman77} and accretion tori
\citep[e.g.][]{christo93}, where physically distinct modes approach
one another in frequency and exchange character. However, we cannot
exclude the possibility  that some modes may have been missed in a 
numerical search of eigenfrequencies.

Thus, our results do not rigorously prove that the GI and MRI branches
do not intersect.  Nevertheless, the continous variation of $\tau$
strongly suggest that unstable modes can transition smoothly from 
MRI to GI and vice versa, especially at low $\beta$.

\subsubsection{Case study} 
In reality, perturbations with a range of $k_x$ will be present for a
given set of disk parameters. Fig. \ref{compare_growth3_Q01d2} show
growth rates as a function of $k_x$ in a disk with
$Q=0.12,\,\beta=20$, where MRI and GI have comparable growth
rates. All perturbations with $k_xH \lesssim 3.5$ grow dynamically
($\gamma\gtrsim 0.1\Omega$, or $\lesssim$ 1.6 orbits).

@@

The dispersion relation $\gamma(k_x)$ deviates from the
pure GI/MRI curves with increasing $k_x$, implying stronger interaction between
magnetic and density perturbations. Comparing pure GI (dashed line)
and the gravitationally-dominated portions of $\gamma(k_x)$ shows that 
inclusion of magnetic field stabilizes high-$k_x$ pure GI. (Note also
the slight decrease in the most unstable $k_x$.) %Since axisymmetric GI
%is unaffected by shear, 
This stabilization is due to magnetic pressure \citep{lizano10}, which
is consistent with stabilizing the small wavelength GI only.

@@

Comparing pure MRI (solid line) and the magnetically-dominated
portions of $\gamma(k_x)$ show that self-gravity increases MRI growth
rates at large $k_x$, but the effect is small. %can be used for code
                                %test 
Note that this 
destabilization is through the linear response, rather than through
the background stratification (which is stabilizing). 
%For the $k_x$ in the GI
%interval, MRI is weakly stabilized by self-gravity.

@@

%implications for simulations -> need to resolve small-scales

%Fig. \ref{compare_growth3_Q01d2} again shows smooth
%transitions between MRI and GI. This means the MRI and GI do indeed
%interact, but

%In particular, modes with $1.5 \lessim
%k_xH\lesssim2$ is neither pure

%demonstrates that mode classification
%becomes ambigious when both GI and MRI are present.

### 2014-04-02 paper/summary.tex
tex updates

\section{Summary and discussion}\label{summary}
In this paper, we have performed explicit stability calculations of
magnetized, self-gravitating disks in the local approximation.

%layered accretion incompatable in massive disks

%why did previous MRI simulations find no difference between SG and
%nonSG? prob because the mri mode has too small vertical wavelength

>>>>

\section{Summary and discussion}\label{summary}
In this paper, we have performed axisymmetric linear stability
calculations of magnetized, self-gravitating disks in the local
approximation. Our models include resistivity and azimuthal fields.  
We have identified regimes under which the magneto-rotational
instability (MRI) is affected by disk self-gravity (SG).  
%what have we extended?

For a vertical field, the requirement for the MRI to operate is that its
vertical wavelength $\lambda \lesssim H$. While $\lambda$ is independent of
the strength of self-gravity $Q$, the disk thickness $H=H(Q)$ %lambda
%depends on omega, which might be
                                %lowered in a massive disk (sub-keplerian)
decreases with increasing SG, which reduces the MRI growth rates.   
%This
%effect becomes noticeable when the disk is marginally stable to GI in
%3D ($Q\sim 0.2$ or $Q_\mathrm{2D}\sim 0.7$).  
Thus, a sufficiently massive disk can surpress the MRI. %but GI may
                                %develop  instead
The MRI is also restricted to larger radial
scales as $Q$ is lowered. This means that the MRI becomes more global
in self-gravitating disks. Finally, the condition $\lambda < H$ 
requries $f(Q)/\sqrt{\beta}$ to be sufficiently small, so weaker
fields are required to permit the MRI with increasing self-gravity.

%need larger boxes 
%This suggests
%that small-scale MRI turbulence may not develop in self-gravitating
%disks.

%correlation

Interestingly, for layered resistivity we do not find layered
perturbations when the disk is massive. This is 
consistent with MRI becoming vertically global with increasing self-gravity.    
For non-self-gravitating disks $\lambda\ll H$, so the MRI can be
restricted to regions of size $L<H$, i.e. an active layer. This is not
possible if $\lambda \sim H$, as found for massive disks. Hence we
find magnetic perturbations penetrate into the high-resistivity dead
zone (e.g. $Q=0.2$ in Fig. \ref{poly_layer}), and there is no distinct
boundary between active and dead layers. This suggests that the
picture of layered accretion  \citep[e.g.][]{fleming03} may not be applicable 
to self-gravitating disks.  %strong magnetization

%SG through 
We find MRI perturbations with radial scale of $\sim H$ can acquire  
density perturbations in a massive but Toomre-stable disk. 
This occurs when the MRI is weak, for example with a strong field or
high resisitivty. We argue in that case $\lambda\simH$, so the density
perturbation caused by the MRI can be enhanced by self-gravity.

%further enhancing density perturbations. 
At this point it is worth mentioning previous non-linear calculations 
performed by \cite{kim03} and \cite{fromang04a,fromang04} in the
context of galactic and circumstellar disks, respectively. These
authors simulated the MRI in self-gravitating disks. They find
self-gravity has little impact on the MRI.

%cite{kim03} and \cite{fromang04}

%Numerical experiments
%varying the field strength and resisitivity indicate this is again a
%result

%GI/MRI: avoided crossings

%azimuthal fields

%\subsection{Implications for non-linear simulations}
%what scaes? resolution
%why don't formang  and kim/ostriker see any difference?
